Associations of environmental and human factors with altered rodent populations in West Africa: A scoping review

### 2021-04-19 reports/manuscript.Rmd
Update manuscript.Rmd

Rodents (Rodentia) are abundant and diverse, representing around 40% of the total number of mammalian species [@american_society_of_mammologists_asm_2021]. Rodents typically demonstrate ‘fast’ life history strategies characterised by early maturation, short generation times, low juvenile and adult survival and high fecundity [@dobson_fast_2007]. There is a degree of between-species heterogeneity, with commensal species and those that are reservoirs of zoonoses having traits consistent with faster life histories (Han et al., 2015). Rodents often compete with humans for food produced in agricultural systems [@fiedler_rodent_1988], with 5-10% of rodent species classed as major pests. Commensal species thrive in human adapted landscapes nesting in houses and obtaining nutrition from human food storage and preparation areas. In sub-Saharan Africa, Mastomys spp. and Arvicanthis spp. are identified as causing a significant burden as agricultural pests [@stenseth_mice_2003]. Crop loss within Afro-Malagasy small-holder farming communities is high, with 15% of pre-harvest crop being lost to rodent pest activity [@swanepoel_systematic_2017].

Another important issue is the transmission of infectious diseases from rodents to humans. Zoonotic diseases (i.e. animal pathogens that impact human health) are transmitted via two pathways – a direct and an indirect pathway. The direct transmission pathway involves transmission of pathogens through rodents biting or scratching a human, or through contamination of food or water sources. The indirect pathway involves rodents acting as amplifying hosts for ectoparasites that themselves are vectors of zoonoses [@meerburg_rodent-borne_2009]. Several rodent-borne pathogens are global in their distribution, including bacteria such as Lyme disease (caused by *Borrelia burgdorferi*) and Leptospirosis (caused by *Leptospira spp.*) and parasites such as Toxoplasmosis (caused by *Toxoplasma gondii*) and Leishmaniasis (caused by *Leishmania spp.*). However, most rodent-borne pathogens have more limited geographic distributions, particularly viral diseases such as Hantavirus pulmonary syndrome (caused by Hantaviridae) concentrated in the Americas and Lassa fever (caused by Arenaviridae) limited to West Africa.

Given the significant economic and public health burden of rodents, monitoring rodent distributions is therefore an ongoing focus of research efforts across multiple scientific disciplines, including (but not limited to) conservation science, development studies and infectious disease epidemiology. The distribution of rodent populations can be monitored via rodent trapping efforts, with traps deployed in habitats expected to contain rodent species of interest.

There is increasing interest in the link between environmental and human factors and rodent populations to land use change and human population change. A recent review of studies included in the PREDICTS database [@hudson_predicts_2014] identified that land use change from primary (i.e. intact forests) to managed and urban systems is associated with a change in rodent assemblages towards species that are more likely to be hosts of zoonoses [@gibb_zoonotic_2020]. *Mastomys natalensis populations* in Tanzania, East Africa are sensitive to seasonal meteorological cycles, in years with below average rainfall and short wet seasons population density is lower [@makundi_reproduction_2007]. Future climate change projections include longer and wetter West African monsoon seasons which could lead to increased zoonotic disease host rodent populations [@akinsanola_projection_2019]. As West Africa contains several known rodent zoonotic reservoirs (e.g. *Mastomys natalensis* and *Arvicanthus niloticus*) and several potential reservoirs of zoonotic diseases (e.g. *Lemniscomys striatus*), this will be the geographic region of interest in this work.

@@

We specifically aim to address the following research questions :

@@

@@

The brand or description of the rodent trap was obtained from each study. For those using multiple devices all types were recorded. Type of trapping method. Effort of trapping

@@

>>>>

Rodents (Rodentia) are abundant and diverse, representing around 40% of the total number of mammalian species [@american_society_of_mammologists_asm_2021]. Rodents typically demonstrate ‘fast’ life history strategies characterised by early maturation, short generation times, low juvenile and adult survival and high fecundity [@dobson_fast_2007]. There is a degree of between-species heterogeneity, with commensal species and those that are reservoirs of zoonoses having traits consistent with faster life histories (Han et al., 2015). Rodents often compete with humans for food produced in agricultural systems [@fiedler_rodent_1988], with 5-10% of rodent species classed as major pests. Commensal species thrive in human adapted landscapes nesting in houses and obtaining nutrition from human food storage and preparation areas. In sub-Saharan Africa, *Mastomys spp.* and *Arvicanthis spp.* are identified as causing a significant burden as agricultural pests [@stenseth_mice_2003]. Crop loss within Afro-Malagasy small-holder farming communities is high, with 15% of pre-harvest crop being lost to rodent pest activity [@swanepoel_systematic_2017].

Another important issue is the transmission of infectious diseases from rodents to humans. Zoonotic diseases (i.e. animal pathogens that impact human health) are transmitted via two pathways – a direct and an indirect pathway. The direct transmission pathway involves transmission of pathogens through a rodent biting or scratching a human, or through contamination of food or water sources. The indirect pathway involves rodents acting as amplifying hosts for ectoparasites that themselves are vectors of zoonoses [@meerburg_rodent-borne_2009]. Several rodent-borne pathogens are global in their distribution, including bacteria such as Lyme disease (caused by *Borrelia burgdorferi*) and Leptospirosis (caused by *Leptospira spp.*) and parasites such as Toxoplasmosis (caused by *Toxoplasma gondii*) and Leishmaniasis (caused by *Leishmania spp.*). However, most rodent-borne pathogens have more limited geographic distributions, particularly viral diseases such as Hantavirus pulmonary syndrome (caused by Hantaviridae) typically concentrated in the Americas and Lassa fever (caused by Arenaviridae) limited to West Africa.

The significant economic and public health burden of rodents has led to the monitoring of rodent distributions being an important focus of research efforts across multiple scientific disciplines, including (but not limited to) conservation science, development studies and infectious disease epidemiology. The distribution of rodent populations can be effectively monitored through rodent trapping, with traps deployed in habitats expected to contain rodent species of interest.

Further, there is increasing interest in the link between environmental and human factors and rodent populations to land use change and human population change. A recent review of studies included in the PREDICTS database [@hudson_predicts_2014] identified that land use change from primary (i.e. intact forests) to managed and urban systems is associated with a change in rodent assemblages towards species that are more likely to be hosts of zoonoses [@gibb_zoonotic_2020]. *Mastomys natalensis* populations in Tanzania, East Africa are sensitive to seasonal meteorological cycles, in years with below average rainfall and short wet seasons population density is lower [@makundi_reproduction_2007] with reduced pest activity as a result. Future climate change projections include longer and wetter West African monsoon seasons which could lead to increased zoonotic disease host rodent populations [@akinsanola_projection_2019]. As West Africa contains several known rodent zoonotic reservoirs (e.g. *Mastomys natalensis* and *Arvicanthus niloticus*) and several potential reservoirs of zoonotic diseases (e.g. *Lemniscomys striatus*), this scoping review will be limited to considering the evidence produced from West Africa.

@@

We specifically aim to address the following research questions:

@@

A scoping review approach was chosen to support a more descriptive and exploratory approach than using a systematic review.

@@

The brand or description of the rodent trap was obtained from each study. For those using multiple devices all types were recorded. The method employed when setting the traps was extracted where reported. Studies that clearly reported setting a transect line were classified as such, for studies reporting using a line but not through a specified habitat trap line was used.

The trapping effort in trap-nights was recorded. A trap-night is the setting of a single trap for a single night. If this was reported at the habitat level or the level at which the study reported rodent captures it was classified as a complete recording of trapping effort. For studies that reported trapping effort at different levels to the capture level it was recorded as incomplete. For studies reporting number of nights trapping occured for or for where the number of traps set was stated but the number of nights not recorded they were classified as incomplete.

@@

The habitat classification scheme a study used was recorded if reported. For studies not using standardised recording of habitat types the explicit description from the study was associated with the trap. For studies reporting multiple habitat types for a single trap, trap-line or trapping grid a higher order classification was recorded to encompass the types.
